Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks

### 2025-05-06 inst/paper/main.tex
add info on subcritical cluster size for cases limit
switch order of contact tracing and isolation

\title{Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks}
\author{Joshua W. Lambert}
\date{}

@@

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005). For highly transmissible pathogens (e.g. $R>3$) with short generation times (e.g. $<3$ days), swift and effective interventions are required to achieve containment. Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}. \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al). Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\

It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that contact tracing and isolation is able to control influenza. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model. In effect the ability to control an epidemic with contact tracing and isolation is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021). The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024). This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.

@@

To assess the efficacy of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}. The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility). The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

@@

There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks. We set this limit to 500 cases.

@@

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5. We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g. genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.

@@

To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of contact tracing and isolation we set up a simulation experiment. We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmissin, and number of initial cases. \\

@@

Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number. A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness. (Figure \ref{fig:prop-outbreak-control-R}). There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of contact tracing and isolation for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once control effectiveness exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}. Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with contact tracing and isolation at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}). \\

The incubation period influences the efficacy of tracing and isolation. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission. The difficulty to control an pandemic outbreak with a short incubation period is shown by the $\sim 40\%$ of simulated H1N1 outbreaks with 100\% control effectiveness that were not contained when $R = 3.5$, while longer incubation period outbreaks have $>80\%$ controlled when 100\% of contacts are traced and isolated (Figure \ref{fig:prop-outbreak-control-R}). Overall, both the transmissibility and incubation period impact the feasibility to control an outbreak. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_reproduction_number.png}
\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an assumed reproduction number ($R$) of either 1.1 or 1.5. This is for a baselines scenario where the number of initial cases is 10, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
\label{fig:prop-outbreak-control-R}
\end{figure}

@@

The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak. The higher the number of cases the less chance that stochastic extinction of a transmission chain will end the whole epidemic. For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.1$ most outbreaks can be controlled without contact tracing and isolation, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}). When the number of cases seeding the outbreak is $>20$ at $R>2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $>60\%$. In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks (Figure \ref{fig:prop-outbreak-control-num-init-cases}).  \\

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_num_init_cases.png}
\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an initial number of seeding infections assumed to be either 5 or 10. This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
\label{fig:prop-outbreak-control-num-init-cases}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_onset_to_isolation.png}
\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan). This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the number of initial infections is 10.}
\label{fig:prop-outbreak-control-onset-to-isolation}
\end{figure}

@@

\item This study has looked at the feasibility and effectiveness of contact tracing and isolation for influenza subtypes that have cause epidemic outbreaks.

@@

\item In modelling the effectiveness of contact tracing and isolation we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref). We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018). We also did not consider zoonotic spillover infections or formite transmission.

@@

\item In this study we model contact tracing and isolation as the sole intervention in containing an influenza outbreak. This is to provide an assessment of its effectiveness for rapidly transmission and potentially highly transmissible subtypes. However, in reality, pandemic preparedness to a pandemic potential influenza would not need to deploy interventions, such as contact tracing individually or sequentially, instead they can be used in combination, such as antivirals (e.g. oseltamivir), rapid tests, contact tracing, and vaccination.

@@

\item This study used a branching process model to simulate individual-level transmission and intervention. The model assumes an infinite population size with no interdependence of contact tracing and isolation between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model. It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing. We suppose that because our model requires isolation sampled for a single individual, our results are more conservative on the ability to control outbreaks, and that using a random network model would indicate an increased effectivness of contact tracing (Juul and Strogatz, 2023). The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025). Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential. If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions. If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence contact tracing and isolation while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
\item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity. Infection severity and fatality risks are of course critical when formulating a pandemic response plan. By understanding the efficacy of contact trace and isolation for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.

@@

\item Contact tracing and isolation complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.

>>>>

\title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
\author{Joshua W. Lambert}
\date{}

@@

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005). For highly transmissible pathogens (e.g. $R>3$) with short generation times (e.g. $<3$ days), swift and effective interventions are required to achieve containment. Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}. \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Isolation and contact tracing is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al). Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\

It has become widely accepted that influenza viruses cannot be controlled with isolation and contact tracing because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model. In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021). The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024). This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.

@@

To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}. The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility). The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

@@

There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks. We set this limit to 500 cases. This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e. larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).

@@

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5. We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g. genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating isolation and contact tracing requires simulating outbreaks that are likely to exponentially spread without control measures.

@@

To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment. We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmissin, and number of initial cases. \\

@@

Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number. A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness. (Figure \ref{fig:prop-outbreak-control-R}). There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once control effectiveness exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}. Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with isolation and contact tracing at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}). \\

The incubation period influences the efficacy of isolating and tracing. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission. The difficulty to control an pandemic outbreak with a short incubation period is shown by the $\sim 40\%$ of simulated H1N1 outbreaks with 100\% control effectiveness that were not contained when $R = 3.5$, while longer incubation period outbreaks have $>80\%$ controlled when 100\% of contacts are traced and isolated (Figure \ref{fig:prop-outbreak-control-R}). Overall, both the transmissibility and incubation period impact the feasibility to control an outbreak. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_reproduction_number.png}
\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an assumed reproduction number ($R$) of either 1.1 or 1.5. This is for a baselines scenario where the number of initial cases is 10, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
\label{fig:prop-outbreak-control-R}
\end{figure}

@@

The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak. The higher the number of cases the less chance that stochastic extinction of a transmission chain will end the whole epidemic. For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.1$ most outbreaks can be controlled without isolation and contact tracing, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}). When the number of cases seeding the outbreak is $>20$ at $R>2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $>60\%$. In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks (Figure \ref{fig:prop-outbreak-control-num-init-cases}).  \\

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_num_init_cases.png}
\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an initial number of seeding infections assumed to be either 5 or 10. This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
\label{fig:prop-outbreak-control-num-init-cases}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_onset_to_isolation.png}
\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan). This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the number of initial infections is 10.}
\label{fig:prop-outbreak-control-onset-to-isolation}
\end{figure}

@@

\item This study has looked at the feasibility and effectiveness of isolation and contact tracing for influenza subtypes that have cause epidemic outbreaks.

@@

\item In modelling the effectiveness of isolation and contact tracing we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref). We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018). We also did not consider zoonotic spillover infections or formite transmission.

@@

\item In this study we model isolation and contact tracing as the sole intervention in containing an influenza outbreak. This is to provide an assessment of its effectiveness for rapidly transmission and potentially highly transmissible subtypes. However, in reality, pandemic preparedness to a pandemic potential influenza would not need to deploy interventions, such as contact tracing individually or sequentially, instead they can be used in combination, such as antivirals (e.g. oseltamivir), rapid tests, contact tracing, and vaccination.

@@

\item This study used a branching process model to simulate individual-level transmission and intervention. The model assumes an infinite population size with no interdependence of isolation and contact tracing between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model. It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing. We suppose that because our model requires isolation sampled for a single individual, our results are more conservative on the ability to control outbreaks, and that using a random network model would indicate an increased effectivness of contact tracing (Juul and Strogatz, 2023). The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025). Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential. If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions. If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence isolation and contact tracing while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
\item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity. Infection severity and fatality risks are of course critical when formulating a pandemic response plan. By understanding the efficacy of isolation and contact trace for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.

@@

\item Isolation and contact tracing complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
